A Plasma Torus Around a Young Low-Mass Star

### 2025-03-18 paper/ms.tex
update derivations

{\it Effective temperature, radius, mass, and spectral type} We adopted
the star's effective temperature and radius measured using the spectral
energy distribution (SED) fitting procedure described by
\cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we performed the fit using the broadband magnitudes from
Gaia DR2, APASS, 2MASS, SDSS, and WISE $W1$ and $W2$.  The resulting
best-fit SED is shown in Figure~\ref{fig:sed}.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
the HIRES spectra against the photometric standards tabulated by
\cite{Bochanski2007}.   This result agrees with the effective
temperature found from the SED fitting \cite{Pecaut2013}.

@@

How much dust is allowed, given the lack of an infrared excess?
Figure~\ref{fig:sed} shows the SED.  While AllWISE \cite{Cutri2014}
yielded a confident W3 detection (9.8$\sigma$), the W4 extraction
yielded only a marginal indication (1.7$\sigma$) of detectable flux.
Similar to other CPVs \cite{Stauffer2017,Bouma2024}, the photometric
uncertainties from WISE W1 and W2 allow at most a $\lesssim$2\% excess
at 3-5\,$\mu$m relative to the stellar photosphere, and a $\lesssim$5\%
excess at 10\,$\mu$m (W3).  To estimate the implied mass bound, we
assume $T_{\rm dust}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and the star as Planck functions,
requiring $F_{\lambda,\star} > f \cdot F_{\lambda,\mathrm{dust}}$, for
$f$ the scaling factor set by the WISE photometric uncertainty.  For
this condition to hold, we find that the bolometric flux ratio $F_{\rm
dust}/F_{\star}$$<$6$\cdot$10$^{-3}$.  Scaling the associated
luminosities ($L = \sigma_{\rm SB} T_{\rm eff}^4 / (4\pi R^2)$)
implies an emitting area for the dust $<$10\% of the star's surface
area.  Converting this constraint to a dust mass requires an
assumption regarding the grain properties; assuming a grain density
$\rho_{\rm d}$=3\,g\,cm$^{-3}$ typical for silicate grains, and a fixed grain
size $a$=1\,$\mu$m \cite{Tanimoto2020,Gunther2022,Koen2023} yields an
upper limit of $M_{\rm dust}\lesssim$10$^{17}$\,g.

By way of comparison, the lower limit follows from requiring the
optical depth $\tau$ to be at least unity.
The optical depth can be written $\tau = n \sigma \ell = N \sigma$,
where $\sigma$ is the cross-section, $n$ is the number density, $\ell$
is the path length, and $N$ is the column density.
For spherical dust grains in the optical, $\sigma = Q_{\rm ext} \pi
a^2$, where $Q_{\rm ext}$ is the extinction efficiency parameter,
tabulated e.g.~by \cite{Croll2014} in their Figure 13.
\cite{Sanderson2023} calculated the relevant cloud mass scaling
assuming a spherical dust clump of size $r$,
and they found
\begin{equation}
  M_{\rm dust} \gtrsim 1.6 \cdot 10^{15}\, {\rm g}\ 
  \left( \frac{\tau}{1} \right)
  \left( \frac{Q_{\rm ext}}{3} \right)^{-1}
  \left( \frac{r}{0.1\,R_\star} \frac{R_\star}{0.34\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}
By way of comparison, prominences of the Sun typically have gas masses
of order $10^{14}\,g$-$10^{16}$\,g \cite{VialEngvold2015}, whereas a
planetesimal of this mass with a bulk density of 1\,g\,cm$^{-3}$ would
have a diameter of order one kilometer.  Halley's comet has a mass of
order $10^{17}$\,g \cite{Rickman1989}, of which of order one part in a
thousand is shed per orbit, most of which inspirals toward the Sun due
to Poynting-Robertson drag.

Broadly, these calculations suggest that if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.

>>>>

{\it Effective temperature, radius, mass, and spectral type}---We adopted
the star's effective temperature and radius measured using the spectral
energy distribution (SED) fitting procedure described by
\cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we performed the fit using the broadband magnitudes from
Gaia DR2, APASS, 2MASS, SDSS, and WISE $W1$ and $W2$.  The resulting
best-fit SED is shown in Figure~\ref{fig:sed}.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
the HIRES spectra against the photometric standards tabulated by
\cite{Bochanski2007}.   This result agrees with the effective
temperature found from the SED fitting \cite{Pecaut2013}.

@@

Regarding the upper limit, Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$), the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and the star as Planck functions, and
require $L_{\rm d} < f L_\star$, where the factor $f$ is set by
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ is an emission efficiency parameter.
Converting this constraint to a dust mass requires an assumption
regarding the grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m \cite{Tanimoto2020,Gunther2022,Koen2023}, and no
self-absorption.  This enables the assumption that $A_{\rm d} = N \pi
a^2$, for $N$ the total number of dust grains.  This in turn yields an
upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

The analogous lower limit follows from requiring the optical depth
from absorption and scattering $\tau$ to be at least unity.  The
optical depth can be written $\tau = n \sigma \ell = N \sigma$, where
$\sigma$ is the cross-section, $n$ is the number density, $\ell$ is
the path length, and $N$ is the column density.  For spherical dust
grains in the optical, $\sigma = Q_{\rm ext} \pi a^2$, where $Q_{\rm
ext}$ is the extinction efficiency parameter, tabulated e.g.~by
\cite{Croll2014} in their Figure 13.  \cite{Sanderson2023} calculated
the relevant cloud mass scaling assuming a spherical dust clump of
size $r$, and they estimated
\begin{equation}
  M_{\rm dust} \gtrsim 2 \cdot 10^{15}\, {\rm g}\ 
  \left( \frac{\tau}{1} \right)
  \left( \frac{Q_{\rm ext}}{3} \right)^{-1}
  \left( \frac{r}{0.1\,R_\star} \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

Three relevant objects for comparison to these limits include solar
prominences, planetesimals, and comets.
Prominences of the Sun typically have gas masses
of order $10^{14}$\,g-$10^{16}$\,g \cite{VialEngvold2015}.
A planetesimal of mass $\approx$10$^{15}$\,g with a bulk density of
1\,g\,cm$^{-3}$ would have a diameter of order one kilometer.
Halley's comet has a mass of order $10^{17}$\,g \cite{Rickman1989}, of
which $\sim$$10^{14}$\,g is shed per orbit, most of which inspirals
toward the Sun due to Poynting-Robertson drag.
Broadly, these calculations suggest that if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.
